A summary on Delta Debugging and its uses

### 2019-03-21 Paper/article.tex
Incorporated Stephans comments and final tweaks.

\author{Moritz C. Laupichler}{Fakultät für Informatik, Karlsruhe Institute of Technology, Germany}{moritz.laupichler@student.kit.edu}{}{}

@@

% Colors
\usepackage{xcolor}
\definecolor{td-green}{rgb}{0,.59,.51}

@@

\begin{abstract}
	Debugging is tedious work for any programmer. This thought sparked the concept of Delta Debugging. In the late 90's and early 2000's Andreas Zeller abstracted a simplistic scenario (``\yd\ my program worked, \td\ it does not. Why?'') into algorithmically tackling debugging. The goal is simple: If every change between \yd\ and \td\ is seen as a Delta, which Deltas are responsible for the failure of \td?

@@

This section outlines the scenario that acts as the base for all following definitions and gives an idea of how Delta Debugging is useful.

@@

Imagine a software development process where the latest changes to the code cause known tests to fail even though an older version of the code passed these exact tests. As the basis of the Delta Debugging principle we need to define a simple abstraction of this point in the development process: We know a state \yd\ that passes the tests and a state \td\ that fails the tests. Furthermore we know all changes that when applied to \yd\ yield \td. \\
The goal of Delta Debugging is to automate extraction of the exact changes to \yd\ that cause the failure in \td.

@@

\definition \defsub{Failure-inducing configuration} A configuration $c$ is called \textbf{failure-inducing} iff 
\[ \forall c' (c \subseteq c' \subseteq \C \rightarrow test(c') \ne \cmark) \] 
holds.

\definition \defsub{Minimal failure-inducing configuration} A failure-inducing configuration $c$ is \textbf{minimal} iff 
\[ \forall c' \subset c: test(c') \ne \xmark \]
holds.\\

@@

There is two major difficulties when approaching a Delta Debugging algorithm: interference and inconsistency. A na\"ive Delta Debugging algorithm called \sdd\ will be refined into the \dd\ algorithm and the \ddp\ algorithm in this chapter in order to describe and solve both difficulties. Finally, isolation of a minimal-failure inducing difference, an alternative approach to DD, is portrayed through the \dddiff\ algorithm.

@@

The na\"ive algorithm \sdd\ recursively splits \C\ into smaller and smaller parts until one failure-inducing change is found. In each recursion step the current set of changes, $c$, is partitioned in two halves, $c_1$ and $c_2$, and both halves are tested. The half that fails the tests is regarded in the next recursion step. The algorithm knows two cases, "Found in $c_1$" and "Found in $c_2$".
\\
\begin{figure}[h!]
\fbox{\begin{minipage}{\textwidth}

\begin{algorithmic}[1]
		\Function{simpledd}{$c: 2^{\C}$}
			\If{$|c| = 1$} \Return $c$ \EndIf
			\State Partition $c$ into two halves $c_1$, $c_2$ so that $c_1 \cap c_2 = \emptyset$
			\If{($test(c_1) = \xmark$)} \Return $simpledd(c_1)$ \Comment("Found in $c_1$") \Else{} \Return $simpledd(c_2)$ \Comment("Found in $c_2$")
			\EndIf
		\EndFunction
\end{algorithmic}
\end{minipage}}
\caption{The na\"ive DD algorithm \sdd}
\label{fig:sdd}
\end{figure}
\\
It becomes evident that this algorithm is not effective because it is only capable of finding single changes that cause failure on their own. But what if multiple changes pass the test individually but only their combination induces failure? \\

\definition \defsub{Interference} Two configurations $c_1$ and $c_2$ \textbf{interfere} iff $test(c_1) = \cmark$, $test(c_2) = \cmark$ but $test(c_1 \cup c_2) = \xmark$. \\

@@

\sdd\ is unable to handle the case of both $c_1$ and $c_2$ passing. A new algorithm \dd\ is created which behaves like \sdd\ but incorporates the case "Interference". In this case the algorithm searches for the failure-inducing change in $c_1$ while keeping $c_2$ applied and vice versa. By combining the results of both searches the interfering changes can be extracted.
\\
\begin{figure}[h!]
\fbox{\begin{minipage}{\textwidth}

$dd = dd_2(\C, \emptyset)$ where \\

\begin{algorithmic}[1]
	\Function{$dd_2$}{$c,r:2^\C$} $: 2^{\C}$
	\State let $c_1,c_2 \subseteq c \text{ with } c_1 \cup c_2 = c, c_1 \cap c_2 = \emptyset, |c_1| \approx |c_2|$
	\State $d_1 = dd_2(c_1, c_2 \cup r)$ \Comment("Search in $c_1$ while leaving $c_2$ applied")
	\State $d_2 = dd_2(c_2, c_1 \cup r)$ \Comment("Search in $c_2$ while leaving $c_1$ applied")\medskip  
	\State \Return $ \begin{cases}
			c & \text{if } |c|=1, \\
			dd_2(c_1,r) & \text{if } test(c_1 \cup r) = \xmark, \text{\Comment("Found in $c_1$")}\\
			dd_2(c_2,r) & \text{if } test(c_2 \cup r) = \xmark, \text{\Comment("Found in $c_2$")}\\
			d_1 \cup d_2 & \text{otherwise } \text{\Comment("Interference")}
		\end{cases}$
	\EndFunction
\end{algorithmic}
\end{minipage}}
\caption{The \dd\ algorithm capable of dealing with interference.}
\label{fig:dd}
\end{figure}
\\

@@

In the real world of code and regression tests this scenario is most likely caused by a combination of changes resulting in syntactical or semantical errors. An inconsistent configuration may also occur when a change is dependent on another change not included in the configuration or when the program execution fails during testing.

\dd\ does not deal with inconsistent configurations. It behaves wrong if the configurations that are tested in the "Found" cases are inconsistent.

@@

The key to solving inconsistency is partitioning $c$ into more than two subsets making the sets of changes smaller. We know that the baseline $\emptyset$ and applying all changes $\C$ are consistent. By applying less changes to \yd\ or removing less changes from \td\ (s. case "Preference" below). the chances of the result being consistent is increased. If the algorithm runs into only inconsistent configurations the number of subsets is increased and the chances of at least one configuration being consistent after is increased. This solution requires the final Delta Debugging algorithm to handle any number $n$ of subsets whereas \dd\ only handles two change sets, $c_1$ and $c_2$ at a time.

\definition \defsub{Complement of configuration} Let $c \subseteq \C$ be a configuration. The configuration $\bar{c}$ is called the \textbf{complement} of $c$. It contains all changes that are contained in the current search but not in $c$.\\

The final Delta Debugging algorithm \ddp\ on the subsets $c_1, \dots, c_n$ knows four cases:\\
\begin{description}
	\item["Found in $c_i$"] If testing any $c_i$ fails it contains a failure-inducing subset. Continue search in $c_i$.
	\item["Interference"] If any $c_i$ and $\bar{c_i}$ both pass they form an interference. Continue as in \dd.
	\item["Preference"] If any $c_i$ is unresolved but $\bar{c_i}$ passes then $c_i$ contains a failure-inducing subset. $c_i$ is then preferred. Continue search in $c_i$ while leaving $\bar{c_i}$ applied to promote consistency.
	\item["Try Again"] If none of the above cases apply then continue the search with $2n$ subsets. The chance of consistent configurations is increased.
\end{description}

\begin{figure}[h!]
\fbox{\begin{minipage}{\textwidth}

$dd^+(c) = dd_3(c, \emptyset,2)$ where \\

\begin{algorithmic}[1]
	\Function{$dd_3$}{$c,r:2^\C, n: \nN$} $: 2^{\C}$
	\State let $c_1,\dots,c_n \subseteq c \text{ with } \bigcup c_i = c, c_i \text{ pairwise disjoint}, \forall i: |c_i| \approx |c|/n$
	\State let $\bar{c_i} = c \setminus (c_i \cup r), t_i = test(c_i \cup r), \bar{t_i} = test(\bar{c_i} \cup r)$
	\State let $c' = c \cap \bigcap \set{\bar{c_i} \mid \bar{t_i} = \xmark}, r' = r \cup \bigcup \set{c_i \mid t_i = \cmark}, n' = min (\mid c' \mid, 2n)$
	\State let $d_i = dd_3(c_i, \bar{c_i} \cup r, 2), \bar{d_i} = dd_3(\bar{c_i},c_i \cup r, 2)$ 
	\State \Return $ \begin{cases}
			c & \text{if } |c|=1, \text{\Comment("Found")}\\
			dd_3(c_i, r, 2) & \text{if } t_i = \xmark \text{ for some i \Comment("Found in $c_i$")}\\
			d_i \cup \bar{d_i} & \text{if } t_i = \cmark \wedge \bar{t_i} = \cmark \text{ for some i \Comment("Interference")}\\
			d_i & \text{if } t_i = \qmark \wedge \bar{t_i} = \cmark \text{ for some i \Comment("Preference")}\\
			dd_3(c',r',n') & \text{if } n < \mid c \mid \text{\Comment("Try Again")}\\
			c' & \text{otherwise \Comment("Nothing Left")}
		\end{cases}$
	\EndFunction
\end{algorithmic}
\end{minipage}}
\caption{The \ddp\ algorithm capable of dealing with unresolved test cases.}
\label{fig:ddp}
\end{figure}

\lemma \ddp\ returns a minimal failure-inducing subset of changes as long as the subset is safe, that is they can be applied to the baseline or removed from \td\ without causing an inconsistency. \ddp\ has linear time complexity.

@@

Another goal that can be achieved by Delta Debugging is the isolation of the minimal failure-inducing difference between \yd\ and \td. Isolation means to replicate this difference through two other configurations \cpass\ and \cfail\ that will not not make any sense from a production standpoint but portray the difference in a minimal and thus simple to understand fashion. This is exactly what the programmer wants when debugging the code: A simplistic reconstruction of what separates passing code from the known failing code in the given context, i.e. the reason why \td\ fails.\\

In more technical terms the goal is to find two configurations \cpass and \cfail, such that $\emptyset \subseteq \cpass \subset \cfail \subseteq \C$, $test(\cpass) = \cmark$ and $test(\cfail) = \xmark$ hold and the difference between the configurations $\Delta = \cfail \setminus \cpass$ is minimal.

\definition \label{def:minimal_difference} \defsub{Minimal failure-inducing difference} Let \cpass\ and \cfail\ be two test cases with $\emptyset \subseteq \cpass \subset \cfail \subseteq \C$. Their difference $\Delta = \cfail \setminus \cpass$ is \textbf{minimal} iff \[ \forall \Delta_i \subset \Delta : test(\cpass \cup \Delta_i) \ne \cmark \wedge test(\cfail \setminus \Delta_i) \ne \xmark \]

@@

\definition \label{def:n-minimal_difference} \defsub{N-minimal failure-inducing difference} Let \cpass\ and \cfail\ be defined as above. Their difference $\Delta = \cfail \setminus \cpass$ is \textbf{n-minimal} iff \[ \forall \Delta_i \subset \Delta : |\Delta_i| \le n \Rightarrow test(\cpass \cup \Delta_i) \ne \cmark \wedge test(\cfail \setminus \Delta_i) \ne \xmark \]

@@

\begin{figure}[t]
	\fbox{
		\begin{minipage}{\textwidth}
		The goal of \dddiff\ is to find $(\cfail, \cpass) = \dddiff(\C)$ such that $\emptyset \subseteq \cpass \subset \cfail \subseteq \C$, $test(\cpass) = \cmark$, $test(\cfail) = \xmark$, and $\Delta = \cfail \setminus \cpass$ is 1-minimal.\\

		The \dddiff\ algorithm is $\dddiff(\C)=dd_4(\emptyset, \C, 2)$ where

				\begin{algorithmic}[1]
					\Function{$dd_4$}{$\cpass,\cfail:2^{\C}$, $n: \nN$}: $2^{\C} \times 2^{\C}$
					\State let $\Delta = \cfail \setminus \cpass$
					\State let $\Delta = \bigcup^{n}_{i=1} \Delta_i$ with all $\Delta_i$ pairwise disjoint and $\forall \Delta_i: |\Delta_i| \approx |\Delta|/n$
					\State let $t_{\cmark, i} = test(\cpass \cup \Delta_i)$ for $i \in \set{1, \dots, n}$ \Comment(``test for larger passing configuration'')
					\State let $t_{\xmark, i} = test(\cfail \setminus \Delta_i)$ for $i \in \set{1, \dots, n}$ \Comment(``test for smaller failing configuration'')
					\State \Return $\begin{cases}
						dd_4(\cpass, \cpass \cup \Delta_i,2) & \text{if } \; t_{\cmark, i} = \xmark \text{\Comment(``reduce to subset'')} \\
						dd_4(\cfail \setminus \Delta_i, \cfail, 2) & \text{if } t_{\xmark, i} = \cmark \text{\Comment(``increase to complement'')}\\
						dd_4(\cpass \cup \Delta_i, \cfail, max(n-1, 2)) & \text{if } t_{\cmark, i} = \cmark \text{\Comment(``increase to subset'')}\\
						dd_4(\cpass, \cfail \setminus \Delta_i, max(n-1,2)) & \text{if } t_{\xmark, i} = \xmark \text{\Comment(``reduce to complement'')}\\
						dd_4(\cpass, \cfail, min(2n, |\Delta|)) & \text{if } n < |\Delta| \text{\Comment(``increase granularity'')} \\
						(\cpass, \cfail) & \text{otherwise \Comment(``done'')} 
					\end{cases}$
					(check the first cases for all $i \in \set{1, \dots, n}$)
					\EndFunction
				\end{algorithmic}
		\end{minipage}
	}
	\caption{The Delta Debugging algorithm for finding two configurations with minimal difference, \dddiff.}
	\label{fig:dddiff}
\end{figure}

@@

\begin{description}
  	\item[``reduce to subset''] If adding any $\Delta_i$ to the current \cpass\ yields a failing configuration, this configuration can be used as a new \cfail\ as it is smaller than the current \cfail. Recurse with a partition granularity of 2.
  	\item[``increase to complement''] If removing any $\Delta_i$ from the current \cfail\ yields a passing configuration, this configuration can be used as a new \cpass\ as it is larger than the current \cpass. Recurse with a partition granularity of 2.
  	\item[``increase to subset''] If adding any $\Delta_i$ to the current \cpass\ yields a passing configuration, this configuration can be used as a new \cpass\ as it is larger than the current \cpass. Recurse with a partition granularity of $n-1$, as the difference changes only by $\Delta_i$ and thus all coarser partitions would result in unresolved tests again.
  	\item[``reduce to complement''] If removing any $\Delta_i$ from the current \cfail\ yields a failing configuration, this configuration can be used as a new \cfail\, as it is smaller than the current \cfail. Recurse with a partition granularity of $n-1$ (s.a.).
  	\item[``increase granularity''] If none of the above cases apply, i.e. the tests all remain unresolved, then the granularity of the partition of $\Delta$ is increased as in \ddp. 
  	\item[``done''] If $n \ge |\Delta|$ the granularity can not be increased further. In that case every $\Delta_i$ only contains one change. If \cpass\ cannot be increased and \cfail\ not be reduced at that point, the difference between them is 1-minimal and they are returned.\\
\end{description}
The full \dddiff\ algorithm is shown in Figure \ref{fig:dddiff}. It is described by Zeller as the successor to \hyperref[fig:ddp]{\ddp} and is recommended to generally replace \ddp\ \cite{Zeller:2002:SIF:506201.506206}. \\

@@

If a certain input is known to cause a program to fail execution Delta Debugging can be used to analyze which part of the program input is responsible for the failure. The \ddp\ algorithm as defined above can be used without any modifications. It is necessary, though, to define certain abstract concepts in the context of input analysis.\\
\begin{itemize}
  	\item We assume a known minimal input for the program at hand that does not cause failure. In many cases this minimal input would be some sort of "empty" input, be it no data, zero values, etc. This minimal input is regarded as the \yd\ state.
  	\item The known failing input is regarded as the \td\ state.
  	\item A change is defined as the smallest possible difference between any two possible inputs, usually a letter, number or other token of data. $\C$ can then be any set of changes that convert the minimal input \yd\ to the failing input \td.
  	\item The $test$ function returns $\xmark$ if exactly the known error occurs on program execution. It returns $\cmark$ if no error occurs and $\qmark$ in all other cases. 
\end{itemize}

@@

It can be of great help for the software developer to be able to automatically determine the specific parts of the program input that cause the failure. The information gathered from the above method can be used by the developer to manually trace the responsible input data to the crash. Considering how a minuscule input change can cause entirely different execution paths this task is tedious if not impossible.
Delta Debugging can help trace the error from the root to the failure during program execution as well, though. Given a test framework that is able to access \hyperref[def:program_state]{program states} during execution, DD is used on the program states of two \hyperref[def:program_run]{program runs} \rpass\ and \rfail\ at significant locations to deduce a \textit{Cause-Effect-Chain (CEC)}.

@@

\definition \label{def:program_run} \defsub{Circumstances and program runs} Let $R$ be the set of all possible circumstance configurations. A run is defined as a configuration of circumstances, so every $r \in R$ is called a run.\\

For our intents and purposes it is sufficient to only consider the program input as the changeable circumstances as we assume the error lies within the program code and occurs independently of operating systems, hardware and other environmental factors.

@@

Consider as an example the above example program \reffail\ that crashes GCC. Applying GCC on $fail.c$ is regarded as \rfail\ and applying GCC to an empty C program as \rpass\ (compiling no code obviously terminates correctly). Andreas Zeller's experimental \textit{HOWCOME} algorithm\footnote{The \textit{HOWCOME} algorithm uses Delta Debugging as described above and incorporates technical issues like the extraction of program states and applying the test and analysis framework \cite{Zeller:2002:ICC:587051.587053}.}
compares the runs at the function \textit{main} shortly after the start of the program, at the function $combine\_instructions$ in the middle of the program run and at the last hit of the function $if\_then\_else\_cond$ shortly before the failure. \textit{HOWCOME} yields a \hyperref[fig:failcec]{Cause-Effect-Chain} that includes the comparisons at those locations.

>>>>

\author{Moritz C. Laupichler}{Department of Informatics, Karlsruhe Institute of Technology, Germany}{moritz.laupichler@student.kit.edu}{}{}

@@

% Colors
\usepackage{xcolor}
\definecolor{td-green}{rgb}{0,.5, 0}

@@

\begin{abstract}
	Debugging is tedious work for any programmer. This thought sparked the concept of Delta Debugging which was first described by Andreas Zeller in the late 90's and early 2000's. A simplistic scenario (``\yd\ my program worked, \td\ it does not. Why?'') was abstracted into algorithmically tackling debugging. The goal is simple: If every change between \yd\ and \td\ is seen as a Delta, which Deltas are responsible for the failure \td?

@@

This section outlines the scenario that acts as the base for all following definitions and gives an idea of the goal of Delta Debugging.

@@

Imagine a software development process where the latest changes to the code cause known tests to fail even though an older version of the code passed these exact tests. As the base of the Delta Debugging principle we need to define a simple abstraction of this point in the development process: We know a state \yd\ that passes the tests and a state \td\ that fails the tests. Furthermore we know all changes that yield \td\ when applied to \yd.\\
The goal of Delta Debugging is to automate extraction of the exact changes to \yd\ that cause the failure in \td.

@@

\definition \defsub{Failure-inducing configuration} A configuration $c$ is called \textbf{failure-inducing} iff 
\[ \forall c' : c \subseteq c' \subseteq \C \Rightarrow test(c') \ne \cmark \] 
holds.

\definition \defsub{Minimal failure-inducing configuration} A failure-inducing configuration $c$ is \textbf{minimal} iff 
\[ \forall c' \subsetneq c: test(c') \ne \xmark \]
holds.\\

@@

There are two major difficulties when approaching a Delta Debugging algorithm: interference and inconsistency. A na\"ive Delta Debugging algorithm called \sdd\ will be refined into the \dd\ algorithm and the \ddp\ algorithm in this chapter in order to describe and solve both difficulties. Finally, isolation of a minimal-failure inducing difference, an alternative approach to DD, is portrayed through the \dddiff\ algorithm.

@@

The na\"ive algorithm \hyperref[fig:sdd]{\sdd} as shown in Figure \ref{fig:sdd} recursively splits \C\ into smaller and smaller parts until one failure-inducing change is found. In each recursion step the current set of changes, $c$, is partitioned into two halves, $c_1$ and $c_2$, and both halves are tested. The half that fails the tests is regarded in the next recursion step. The algorithm knows two cases, "Found in $c_1$" and "Found in $c_2$".
\\
\begin{figure}[h!]
\fbox{\begin{minipage}{\textwidth}
$simpledd(\C) = dd_1(\C)$ where \\

\begin{algorithmic}[1]
		\Function{$dd_1$}{$c: 2^{\C}$}
			\If{$|c| = 1$} \Return $c$ \EndIf
			\State Partition $c$ arbitrarily into two parts $c_1$, $c_2$ such that $|c_1| \approx |c_2| \approx |c|/2$
			\State \Return $\left \{ \text{\begin{tabular}{l l r}
				$dd_1(c_1)$ & if $test(c_1) = \xmark$ & \Comment("Found in $c_1$") \\
				$dd_1(c_2)$ & otherwise & \Comment("Found in $c_2$")
			\end{tabular}}\right.$
		\EndFunction
\end{algorithmic}
\end{minipage}}
\caption{The na\"ive DD algorithm \sdd}
\label{fig:sdd}
\end{figure}
\\

\begin{samepage}

It becomes evident that this algorithm is not effective because it is only capable of finding single changes that cause failure on their own. But what if multiple changes pass the test individually but only their combination induces failure? \\

\definition \defsub{Interference} Two configurations $c_1$ and $c_2$ \textbf{interfere} iff $test(c_1) = \cmark$, $test(c_2) = \cmark$ but $test(c_1 \cup c_2) = \xmark$. \\
\end{samepage}

@@

\sdd\ is unable to handle the case of both $c_1$ and $c_2$ passing i.e. interfering configurations. A new algorithm \dd\ is created which behaves like \sdd\ but incorporates the case "Interference" as shown in Figure \ref{fig:dd}. In this case the algorithm searches for the failure-inducing change in $c_1$ while keeping $c_2$ applied and vice versa. By combining the results of both searches the interfering changes can be extracted.
\\
\begin{figure}[h!]
\fbox{\begin{minipage}{\textwidth}

$dd(\C) = dd_2(\C, \emptyset)$ where \\

\begin{algorithmic}[1]
	\Function{$dd_2$}{$c,r:2^\C$} $: 2^{\C}$
	\State Partition $c$ arbitrarily into two parts $c_1$, $c_2$ such that $|c_1| \approx |c_2| \approx |c|/2$
	\State $d_1 = dd_2(c_1, c_2 \cup r)$ \Comment("Search in $c_1$ while leaving $c_2$ applied")
	\State $d_2 = dd_2(c_2, c_1 \cup r)$ \Comment("Search in $c_2$ while leaving $c_1$ applied")\medskip  
	\State \Return $\left \{ \text{\begin{tabular}{l l r}
			$c$ & if $|c|=1$ & \Comment("Found") \\
			$dd_2(c_1,r)$ & if $test(c_1 \cup r) = \xmark$ & \Comment("Found in $c_1$")\\
			$dd_2(c_2,r)$ & if $test(c_2 \cup r) = \xmark$ & \Comment("Found in $c_2$")\\
			$d_1 \cup d_2$ & otherwise & \Comment("Interference")
		\end{tabular}}
			\right.$
	\EndFunction
\end{algorithmic}
\end{minipage}}
\caption{The \dd\ algorithm capable of dealing with interference.}
\label{fig:dd}
\end{figure}
\\

@@

In the real world of code and regression tests this scenario is most likely caused by a combination of changes resulting in syntactical or semantic errors. An inconsistent configuration may also occur when a change is dependent on another change, that is not included in the configuration, or when the program execution fails during testing.

\dd\ does not deal with inconsistent configurations. It delivers wrong results if the configurations that are tested in the "Found" cases are inconsistent.

@@

The key to solving inconsistency is partitioning $c$ into more than two subsets making the sets of changes smaller. We know that the baseline $\emptyset$ and applying all changes $\C$ are consistent. By applying less changes to \yd\ or removing less changes from \td\ (s. case "Preference" below) the chances of the result being consistent are increased. If the algorithm runs into only inconsistent configurations the number of subsets is increased, thus increasing the chances of at least one configuration being consistent after. This solution requires the new Delta Debugging algorithm to handle any number $n$ of subsets whereas \dd\ only handles two change sets, $c_1$ and $c_2$, at a time.

\definition \defsub{Complement of configuration} Let $c \subseteq \C$ be a configuration that is currently being searched for failure-inducing changes. Let $r \subseteq \C$ be the changes that remain applied during the search. For a subset $a \subseteq c$ the \textbf{complement} is $\bar{a} = c \setminus (a \cup r)$.\\

The new Delta Debugging algorithm \ddp\ on the subsets $c_1, \dots, c_n$ as shown in Figure \ref{fig:ddp} knows four main cases:\\
\begin{description}
	\item["Found in $c_i$"] If testing any $c_i$ fails it contains a failure-inducing subset. Continue search in $c_i$.
	\item["Interference"] If any $c_i$ and $\bar{c_i}$ both pass they form an interference. Continue as in \dd.
	\item["Preference"] If any $c_i$ is unresolved but $\bar{c_i}$ passes then $c_i$ contains a failure-inducing subset. $c_i$ is then preferred. Continue search in $c_i$ while leaving $\bar{c_i}$ applied to promote consistency.
	\item["Try Again"] If none of the above cases apply then continue the search with $2n$ subsets. The chances of consistent configurations are increased.
\end{description}

\begin{figure}[H]
\fbox{\begin{minipage}{\textwidth}

$dd^+(\C) = dd_3(\C, \emptyset,2)$ where \\

\begin{algorithmic}[1]
	\Function{$dd_3$}{$c,r:2^\C, n: \nN$} $: 2^{\C}$
	\State Partition $c$ arbitrarily into $n$ parts $c_1, \dots, c_n$ such that $\forall i: |c_i| \approx |c|/n$
	\State let $\bar{c_i} = c \setminus (c_i \cup r), t_i = test(c_i \cup r), \bar{t_i} = test(\bar{c_i} \cup r)$
	\State let $c' = c \cap \bigcap \set{\bar{c_i} \mid \bar{t_i} = \xmark}, r' = r \cup \bigcup \set{c_i \mid t_i = \cmark}, n' = min (|c'|, 2n)$
	\State let $d_i = dd_3(c_i, \bar{c_i} \cup r, 2), \bar{d_i} = dd_3(\bar{c_i},c_i \cup r, 2)$
	\State \Return $\left \{ \text{\begin{tabular}{l l r}
		$c$ & if |c|=1 & \Comment("Found")\\
		$dd_3(c_i, r, 2)$ & if $t_i = \xmark$ for some $i$ & \Comment("Found in $c_i$")\\
		$d_i \cup \bar{d_i}$ & if $t_i = \cmark \wedge \bar{t_i} = \cmark$ for some $i$ & \Comment("Interference")\\
		$d_i$ & if $t_i = \qmark \wedge \bar{t_i} = \cmark$ for some $i$ & \Comment("Preference")\\
		$dd_3(c',r',n')$ & if $n < |c|$ & \Comment("Try Again")\\
		$c'$ & otherwise & \Comment("Nothing Left")
	\end{tabular}}\right.$
	\EndFunction
\end{algorithmic}
\end{minipage}}
\caption{The \ddp\ algorithm capable of dealing with unresolved test cases.}
\label{fig:ddp}
\end{figure}

\lemma \ddp\ returns a minimal failure-inducing subset of changes as long as the subset is safe, i.e. they can be applied to \yd\ or removed from \td\ without causing an inconsistency. \ddp\ has linear time complexity.

@@

Another goal that can be achieved by Delta Debugging is the isolation of the minimal failure-inducing difference between \yd\ and \td. Isolation means to replicate this difference through two other configurations \cpass\ and \cfail\ that will not make any sense from a production standpoint but portray the difference in a minimal and thus simple to understand fashion. This is exactly what the programmer wants when debugging the code: A simplistic reconstruction of what separates passing code from the known failing code in the given context, i.e. the reason why \td\ fails.\\

In more technical terms the goal is to find two configurations \cpass\ and \cfail, such that $\emptyset \subseteq \cpass \subsetneq \cfail \subseteq \C$, $test(\cpass) = \cmark$ and $test(\cfail) = \xmark$ hold and the difference between the configurations $\Delta = \cfail \setminus \cpass$ is minimal.

\definition \label{def:minimal_difference} \defsub{Minimal failure-inducing difference} Let \cpass\ and \cfail\ be two test cases with $\emptyset \subseteq \cpass \subsetneq \cfail \subseteq \C$. Their difference $\Delta = \cfail \setminus \cpass$ is \textbf{minimal} iff \[ \forall \Delta' \subsetneq \Delta : test(\cpass \cup \Delta') \ne \cmark \text{ and } test(\cfail \setminus \Delta') \ne \xmark \]

@@

\definition \label{def:n-minimal_difference} \defsub{N-minimal failure-inducing difference} Let \cpass\ and \cfail\ be defined as above. Their difference $\Delta = \cfail \setminus \cpass$ is \textbf{n-minimal} iff \[ \forall \Delta' \subsetneq \Delta : |\Delta'| \le n \Rightarrow (test(\cpass \cup \Delta') \ne \cmark \text{ and } test(\cfail \setminus \Delta') \ne \xmark) \]

@@

@@

\begin{description}
  	\item[``reduce to subset''] If adding any $\Delta_i$ to the current \cpass\ yields a failing configuration, this configuration can be used as a new \cfail\ as it is smaller than the current \cfail. Recurse with a partition granularity of 2.
  	\item[``increase to complement''] If removing any $\Delta_i$ from the current \cfail\ yields a passing configuration, this configuration can be used as a new \cpass\ as it is larger than the current \cpass. Recurse with a partition granularity of 2.
  	\item[``increase to subset''] If adding any $\Delta_i$ to the current \cpass\ yields a passing configuration, this configuration can be used as a new \cpass\ as it is larger than the current \cpass. Recurse with a partition granularity of $n-1$, as the difference changes only by $\Delta_i$ and thus all coarser partitions would result in unresolved tests again.
  	\item[``reduce to complement''] If removing any $\Delta_i$ from the current \cfail\ yields a failing configuration, this configuration can be used as a new \cfail\, as it is smaller than the current \cfail. Recurse with a partition granularity of $n-1$ (s.a.).
  	\item[``increase granularity''] If none of the above cases apply, i.e. the tests all remain unresolved, then the granularity of the partition of $\Delta$ is increased as in \ddp. 
  	\item[``done''] If $n \ge |\Delta|$ the granularity can not be increased further. In that case every $\Delta_i$ only contains one change. If \cpass\ cannot be increased and \cfail\ not be reduced at that point, the difference between them is 1-minimal and they are returned.\\
\end{description}
The full \dddiff\ algorithm is shown in Figure \ref{fig:dddiff}. It is described by Zeller as the successor to \hyperref[fig:ddp]{\ddp} and is recommended to generally replace \ddp\ as it is superior in performance and result quality \cite{Zeller:2002:SIF:506201.506206}. \\

\begin{figure}[H]
	\fbox{
		\begin{minipage}{\textwidth}

		$\dddiff(\C)=dd_4(\emptyset, \C, 2)$ where \\

				\begin{algorithmic}[1]
					\Function{$dd_4$}{$\cpass,\cfail:2^{\C}$, $n: \nN$}: $2^{\C} \times 2^{\C}$
					\State let $\Delta = \cfail \setminus \cpass$
					\State Partition $\Delta$ arbitrarily into $n$ parts $\Delta_1, \dots, \Delta_n$ such that $\forall i: |\Delta_i| \approx |\Delta|/n$
					\State let $t_{\cmark, i} = test(\cpass \cup \Delta_i)$ for $i \in \set{1, \dots, n}$ \Comment(``test for larger passing configuration'')
					\State let $t_{\xmark, i} = test(\cfail \setminus \Delta_i)$ for $i \in \set{1, \dots, n}$ \Comment(``test for smaller failing configuration'')
					\State \Return $\left \{ \text{\begin{tabular}{l l r}
						$dd_4(\cpass, \cpass \cup \Delta_i,2)$ & if $t_{\cmark, i} = \xmark$ & \Comment(``reduce to subset'') \\
						$dd_4(\cfail \setminus \Delta_i, \cfail, 2)$ & if $t_{\xmark, i} = \cmark$ & \Comment(``increase to complement'')\\
						$dd_4(\cpass \cup \Delta_i, \cfail, max(n-1, 2))$ & if $t_{\cmark, i} = \cmark$ & \Comment(``increase to subset'')\\
						$dd_4(\cpass, \cfail \setminus \Delta_i, max(n-1,2))$ & if $t_{\xmark, i} = \xmark$ & \Comment(``reduce to complement'')\\
						$dd_4(\cpass, \cfail, min(2n, |\Delta|))$ & if $n < |\Delta|$ & \Comment(``increase granularity'')\\
						(\cpass, \cfail) & otherwise & \Comment(``done'')
					\end{tabular}}\right.$
					(check the first cases for all $i \in \set{1, \dots, n}$)
					\EndFunction
				\end{algorithmic}
		\end{minipage}
	}
	\caption{The Delta Debugging algorithm for finding two configurations with minimal difference, \dddiff. It finds $(\cfail, \cpass) = \dddiff(\C)$ such that $\emptyset \subseteq \cpass \subsetneq \cfail \subseteq \C$, $test(\cpass) = \cmark$, $test(\cfail) = \xmark$, and $\Delta = \cfail \setminus \cpass$ is 1-minimal.}
	\label{fig:dddiff}
\end{figure}

@@

If a certain input is known to cause a program to fail execution then Delta Debugging can be used to analyse which part of the program input is responsible for the failure. The \ddp\ algorithm as defined above can be used without any modifications. It is necessary, though, to define certain abstract concepts in the context of input analysis.\\
\begin{itemize}
  	\item We assume a known minimal input for the program at hand that does not cause failure. In many cases this minimal input would be some sort of "empty" input, be it no data, zero values, etc. This minimal input is regarded as the \yd\ state.
  	\item The known failing input is regarded as the \td\ state.
  	\item A change is defined as the smallest possible difference between any two possible inputs, usually a letter, number or other token of data. $\C$ can then be any set of changes that convert the minimal input \yd\ to the failing input \td.
  	\item The $test$ function returns $\xmark$ if exactly the known error occurs on program execution. It returns $\cmark$ if no error occurs and $\qmark$ in all other cases. 
\end{itemize}

@@

It can be of great help for the software developer to automatically determine the specific parts of the program input that cause the failure. The information gathered from the above method can be used by the developer to manually trace the responsible input data to the crash. Considering how a minuscule input change can cause entirely different execution paths this task is tedious if not impossible.
Delta Debugging can help trace the error from the root to the failure during program execution as well, though. Given a debugging framework that is able to access \hyperref[def:program_state]{program states} during execution, DD is used on the program states of two \hyperref[def:program_run]{program runs} \rpass\ and \rfail\ at significant locations to deduce a \textit{Cause-Effect-Chain (CEC)}.

@@

\definition \label{def:program_run} \defsub{Circumstances and program runs} Let $R$ be the set of all possible circumstance configurations. A (program) run is defined as a configuration of circumstances, so every $r \in R$ is called a run.\\

For our intents and purposes it is sufficient to only consider the program input as the changeable circumstances as we assume the error lies within the program code and occurs independently of operating systems, hardware and other environmental factors. Thus, in our case a program run is equivalent to just the specific input that induces the program run.

@@

Consider as an example the above example program \reffail\ that crashes GCC. Applying GCC on $fail.c$ is regarded as \rfail\ and applying GCC to an empty C program as \rpass\ (compiling no code obviously terminates correctly). Andreas Zeller's experimental \textit{HOWCOME} algorithm\footnote{The \textit{HOWCOME} algorithm uses Delta Debugging as described above and incorporates technical issues like the extraction of program states and applying the debugging framework \cite{Zeller:2002:ICC:587051.587053}.}
compares the runs at the function \textit{main} shortly after the start of the program, at the function $combine\_instructions$ and at each hit of the function $if\_then\_else\_cond$. \textit{HOWCOME} yields a \hyperref[fig:failcec]{Cause-Effect-Chain} that includes the comparisons at those locations as shown in Figure \ref{fig:failcec}.
